---
title: "Parser Combinators"
author: "S Wotherspoon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Parsers

A parser is a function that accepts a string as input and returns
`NULL` on failure or a list with elements

* `input` the unconsumed input, and
* `result` an object representing the result of the parse

if the string satisfies the grammar recognized by the parser.

A parser combinator is a higher order function that combines one or
more simpler parsers to create a parser that recognizes a more complex
grammar.

## Primitive Parsers

The `pLiteral` function generates a primitive parser that recognizes a
literal string.
```{r}
library(Combin8R)
pDog <- pLiteral("Tag","dog")
pDog("monkey")
pDog("dog")
```
The matched literal is returned as the value of the parse, as an S3
object with a class defined by the `tag` argument and inheriting from
`pLiteral`,
```{r}
unclass(pDog("dog")$result)
```
and print methods are defined for the major classes to simplify the
display of the result
```{r}
pDog("dog")$result
```

The `pRegex` function generates a primitive parser that accepts a
grammar defined by a regular expression.  The matched string and any
captured groups from the regular expression are returned as the value
of the parse
```{r}
pLabel <- pRegex("Label","label (\\d+)")
pLabel("label 7")
```

The `tag` argument can be a function that is used to construct the
object representing the value of the parse
```{r}
pLabel <- pRegex(function(value) structure(list(value=as.numeric(value[[2]])),class=c("Label","pRegex")),
                 "label (\\d+)")
pLabel("label 7")
```

## Combinators

There are four main combinators

* `pAlt` creates a parser that accepts input accepted by any one of a
  number of simpler parsers

* `pSeq` creates a parser that accepts input accepted by a number of
  simpler parsers applied in sequence

* `pMany` creates a parser that accepts input for which a simpler
  parser succeeds zero or more times in succession

* `pSome` creates a parser that accepts input for which a simpler
  parser succeeds one or more times in succession



# Simple Logo

Consider the subset of the Logo language consisting of the constructs
repeat, forward, left and right, so that a possible program is given
by
```{r}
program <- "repeat 10 [right 36 repeat 5 [forward 54 right 72]]"
```

## Simple Parser

We can write a parser for the Logo subset as follows
```{r}
pInteger <- pRegex("Integer","\\d+")
pSpaces <- pRegex("Spaces","\\s*")
pSpaces1 <- pRegex("Spaces1","\\s+")
pCommands <-
  pSome("Commands",
    pSeq("CommandWhite",
      pAlt("Command",
        pSeq("forwardCmd",pLiteral("forward"),pSpaces1,pInteger),
        pSeq("rightCmd",pLiteral("right"),pSpaces1,pInteger),
        pSeq("leftCmd",pLiteral("left"),pSpaces1,pInteger),
        pSeq("repeatCmd",pLiteral("repeat"),pSpaces1,pInteger,pSpaces1,pBlock)
      ),
      pSpaces))
pBlock <- pSeq("Block",pLiteral("["),pSpaces,pCommands,pLiteral("]"))
```

This parses the program text and produces an abstract syntax tree (AST)
```{r}
p <- pCommands(program)
p
```


The ``formatAST` generic is used to create a string representation of
the AST and is used to implement the print methods for the AST.  The
printed representation of the AST can be simplified by writing
specific methods for the result classes
```{r}
formatAST.Integer <- function(x,indent,...)
  as.character(x$value)
formatAST.forwardCmd <- function(x,indent,...)
  paste("forward",formatAST(x$value[[3]]))
formatAST.rightCmd <- function(x,indent,...)
  paste("right",formatAST(x$value[[3]]))
formatAST.leftCmd <- function(x,indent,...)
  paste("left",formatAST(x$value[[3]]))
formatAST.repeatCmd <- function(x,indent,...)
  paste("repeat",formatAST(x$value[[3]]),formatAST(x$value[[5]]))
formatAST.Command <- function(x,indent,...)
  formatAST(x$value)
formatAST.CommandWhite <- function(x,indent,...)
  formatAST(x$value[[1]])
formatAST.Commands <- function(x,indent,...)
  paste(sapply(x$value,formatAST),collapse=" ")
formatAST.Block <- function(x,indent,...)
  paste("[",formatAST(x$value[[3]]),"]",sep="")
```

The AST resulting from the parse does not change, only its printed
representation
```{r}
p
```


## Simplified AST

The AST constructed by the simple parser is more complicated than
strictly necessary because it contains nodes for purely syntactic
elements of the language such as whitespace.  These elements can be
removed if we construct the AST ourselves
```{r}
mkprm <- function(value) structure(list(value[[1]]$value,value[[3]]),class="prm")
mkrpt <- function(value) structure(list(value[[3]],value[[5]]),class="rpt")
mkblk <- function(value) structure(value[[3]],class="blk")
pInteger <- pRegex(as.numeric,"\\d+")
pSpaces <- pRegex("Spaces","\\s*")
pSpaces1 <- pRegex("Spaces1","\\s+")
pCommands <-
  pSome(function(value) value,
        pSeq(function(value) value[[1]],
             pAlt(function(value) value,
                  pSeq(mkprm,pLiteral("forward"),pSpaces1,pInteger),
                  pSeq(mkprm,pLiteral("right"),pSpaces1,pInteger),
                  pSeq(mkprm,pLiteral("left"),pSpaces1,pInteger),
                  pSeq(mkrpt,pLiteral("repeat"),pSpaces1,pInteger,pSpaces1,pBlock)
             ),
             pSpaces))
pBlock <- pSeq(mkblk,pLiteral("["),pSpaces,pCommands,pLiteral("]"))
```

The AST now takes the form
```{r}
p <- pCommands(program)
p
```

As before, the print representation can be simplified by by defining
appropriate methods for `formatAST`
```{r}
formatAST.prm <- function(x,indent,...)
  paste(x[[1]],x[[2]],collapse=" ")
print.prm <- function(x,indent=0,...)
  cat(formatAST(x,indent))
formatAST.rpt <- function(x,indent,...)
  paste("repeat",x[[1]],formatAST(x[[2]]),collapse=" ")
print.rpt <- function(x,indent=0,...)
  cat(formatAST(x,indent))
formatAST.blk <- function(x,indent,...)
  paste("[",paste(sapply(x,formatAST),collapse=" "),"]",sep="")
print.blk <- function(x,indent=0,...)
  cat(formatAST(x,indent))
```
so now
```{r}
p
```


## Compiler

Alternately, a compiler for the Logo subset can be written by
translating Logo constructs directly to R expressions
```{r}
mkiter <- local({n <- 0
  function() {
    n <<- n+1
    as.name(paste("k",n,sep=""))
  }})
cmpfwd <- function(value)
  substitute({
      x0 <- x1
      x1 <- r*c(cos(theta),sin(theta))
      segments(x0[1],x0[2],x1[1],x1[2])
    },list(r=value[[3]]))
cmprgt <- function(value)
  substitute(theta <- theta + pi/180*a,list(a=value[[3]]))
cmplft <- function(value)
  substitute(theta <- theta - pi/180*a,list(a=value[[3]]))
cmprpt <- function(value)
  substitute(for(k in 1:n) b,
    list(k=mkiter(),n=value[[3]],b=value[[5]]))
cmpblk <- function(value)
  if(length(value)>1) as.call(c(as.name("{"),value[[3]])) else value
pInteger <- pRegex(as.numeric,"\\d+")
pSpaces <- pRegex("Spaces","\\s*")
pSpaces1 <- pRegex("Spaces1","\\s+")
pCommands <-
  pSome(function(value) value,
        pSeq(function(value) value[[1]],
             pAlt(function(value) value,
                  pSeq(cmpfwd,pLiteral("forward"),pSpaces1,pInteger),
                  pSeq(cmprgt,pLiteral("right"),pSpaces1,pInteger),
                  pSeq(cmplft,pLiteral("left"),pSpaces1,pInteger),
                  pSeq(cmprpt,pLiteral("repeat"),pSpaces1,pInteger,pSpaces1,pBlock)
             ),
             pSpaces))
pBlock <- pSeq(cmpblk,pLiteral("["),pSpaces,pCommands,pLiteral("]"))
pLogo <- function(input,xlim=c(-100,100),ylim=c(-100,100)) {
  p <- pCommands(input)
  if(!is.null(p)) {
    body <- as.call(c(as.name("{"),
                      quote(plot.new()),
                      substitute(plot.window(xlim,ylim)),
                      quote(x1 <- c(0,0)),
                      quote(theta <- 0),
                      p$result))
    list(input=p$input,
         result=substitute(local(body),list(body=body)))
    }
}
```

The result of the parser is now R code
```{r}
p <- pLogo(program)
p
```
and evaluating this code executes the Logo program
```{r}
eval(p$result)
```
